---
title: "Joint-species distribution modelling with Hmsc (S5)"
output: html_notebook
---
This script is a template to guide the application of joint species distribution modelling (jSDM) using the package Hmsc. This code is based primarily off the scripts provided in the book *Joint Species Distribution Modelling: With Applications in R* by Nerea Abrego and Otso Ovaskainen and by Dr. Fabrice Stephenson (University of Waikato).

The primary purpose here is to predict the past, present, and future distributions of epi-megabenthic taxa on a New Zealand seamount as they relate to previous bottom trawl fishing impacts. The main data utilised here was derived from seafloor images collected by National Institute of Water and Atmospheric Research (NIWA) researchers using a towed camera five times between 2001 and 2020. These images were analysed by myself, Savannah Goode (VUW/NIWA), to collect species counts for use in jSDM. Commercial fisheries data was supplied by Ministry for Primary Industries and analysed by Dr. Malcolm Clark (NIWA) and myself to estimate the fine-scale fishing footprint within the study area. Multibeam data was collected and processed by Arne Pallentin.

The script is separated into five steps (with this being the third):

1. Setting model structure and fitting the model.

2. Examining MCMC convergence.

3. Evaluating model fit and comparing models.

4. Exploring parameter estimates.

5. **Making predictions.**

***

#### **Read in data**

Set the working directory to the location of all the relevant files. Abrego and Ovaskainen also recommend creating the subdirectories 'data' and 'models' within the main directory. If running on a different computer or files moved, the wd parameter will have to be changed to match your file structure.
```{r "setup", include = TRUE, echo = FALSE, warning=FALSE}
library(knitr)
# change to match computer directory
opts_knit$set(root.dir = "/Users/savannahgoode/OneDrive - Victoria University of Wellington - STAFF/R/chapter 4")
#opts_knit$set(root.dir = "C:/Users/goodes/OneDrive - Victoria University of Wellington - STAFF/R/chapter 4")
opts_chunk$set(echo = FALSE)
```

```{r}
# change to match computer directory
wd = "/Users/savannahgoode/OneDrive - Victoria University of Wellington - STAFF/R/chapter 4"
#wd = "C:/Users/goodes/OneDrive - Victoria University of Wellington - STAFF/R/chapter 4"
setwd(wd)
localDir = "."
data.directory = file.path(localDir, "Data")
model.directory = file.path(localDir, "Models")
getwd()
```

Load Hmsc library and set seed so results are reproducible.
```{r, message=FALSE, warning=FALSE}
library(Hmsc)
set.seed(3)
```

First read in model objects. Make sure to load the *fitted* models with the highest thin values to get the most reliable results.
```{r}
# set parameters to load model
# change these if needed
nChains = 5
samples = 200
thin = 100
filename=file.path(model.directory,
                   paste0("model_chains_",as.character(nChains),
                          "_samples_",as.character(samples),"_thin_",as.character(thin),
                          ".Rdata",sep=""))
load(filename)

# make sure the right models were loaded
nm <- length(models)
for(i in 1:nm){
  print(paste0("model = ",modelnames[i]))
  print(models[[i]])
}
```
***

#### **Make gradient plots**
Make gradient plots to visualise how the communities vary across environmental covariates.

First plot how taxon richness varies with each environmental covariate.
```{r, warning=FALSE, fig.height = 5, fig.width = 6, message=FALSE}
m <- models[[1]]
sp <- colnames(m$Y)
covariates <- all.vars(m$XFormula)
#covariates[6] <- "(log.nTows * tsTow)"
#covariates <- covariates[c(1:3,5:6)]

par(mfrow = c(2,3))
for(k in 1:(length(covariates))){
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate, non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE, predictEtaMean = TRUE)  
    plotGradient(m, Gradient, pred=predY, measure="S", 
                 showData = TRUE, showPosteriorSupport = FALSE)
    }
```

Then export as PDF.
```{r,warning=FALSE, echo=T}
#filename = paste("Figures/Pres-Abs_predictions_Richness_type1.pdf")
filename =  paste("Figures/Pres-Abs_predictions_Richness_type2.pdf")
pdf(file = filename, height = 4.5, width = 4.5)
  
for(k in 1:(length(covariates))){
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate)
   # Gradient = constructGradient(m,focalVariable = covariate, non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)  
    plotGradient(m, Gradient, pred=predY, yshow = 0, measure="S", showData = TRUE, 
                 main = paste0(modelnames[[1]],": summed response"))
    }

dev.off()
```

Let's examine how individual taxa vary across the different environmental gradients. Here we will just plot one taxon (Comatulid crinoids) as an example then save all results to a PDF.
```{r, warning=FALSE}
# presence-absence
par(mfrow = c(2,3))
for(k in 1:(length(covariates))){
    m = models[[1]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate, non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
    plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                 showData = TRUE, index = 13, showPosteriorSupport = FALSE)
}

# log abundance
par(mfrow = c(2,3))
for(k in 1:(length(covariates))){
    m = models[[2]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate, 
                                 non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE, predictEtaMean = TRUE)
    plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                 showData = TRUE, index = 13, showPosteriorSupport = FALSE)
    }
```

Now we can make all the same plots for each individual taxon and save the results to a pdf. 
```{r, warning=FALSE,echo =TRUE}
filename = paste("Figures/Pres-Abs_predictions_byTaxon_type2.pdf")
#filename =  paste("Figures/Pres-Abs_predictions_byTaxon_type1.pdf") #change if needed depending on pred type
pdf(file = filename, height = 4.5, width = 4.5)

#par(mfrow = c(2,5))
for(k in 1:length(covariates)){
    m = models[[1]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate)
   # Gradient = constructGradient(m,focalVariable = covariate, non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
    for(l in 1:length(sp)){
        plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                     showData = TRUE, index = l,
                     showPosteriorSupport = TRUE)
    }
}

filename =  paste("Figures/LogAbun_predictions_byTaxon_type2.pdf") 
#filename =  paste("Figures/LogAbun_predictions_byTaxon_type1.pdf") #change if needed
pdf(file = filename, height = 4.5, width = 4.5)

for(k in 1:length(covariates)){
    m = models[[2]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m, focalVariable = covariate)
  #  Gradient = constructGradient(m,focalVariable = covariate,non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
    for(l in 1:length(sp)){
        plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                     showData = TRUE, index = l,
                     showPosteriorSupport = TRUE)
    }
}
dev.off()
```

We can also plot how community-weighted means of each trait vary with the environmental covariates.
```{r, warning=FALSE, echo=TRUE}
# presence-absence
filename =  paste("Figures/Pres-Abs_predictions_byTrait.pdf")
pdf(file = filename, height = 4.5, width = 4.5)

m = models[[1]]
nt <- m$nt
tr <- colnames(m$Tr)

for(k in 1:length(covariates)){
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate,non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
  #  par(mfrow = c(2,2))
    for(l in 2:nt){
        plotGradient(m, Gradient, pred=predY, yshow = 0, measure="T", 
                     showData = TRUE, index = l,
                     showPosteriorSupport = TRUE)
    }
}

# abundance
filename =  paste("Figures/LogAbun_predictions_byTrait.pdf")
pdf(file = filename, height = 4.5, width = 4.5)

m = models[[2]]
nt <- m$nt
tr <- colnames(m$Tr)

for(k in 1:length(covariates)){
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate,
                                 non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
  #  par(mfrow = c(2,2))
    for(l in 2:nt){
        plotGradient(m, Gradient, pred=predY, yshow = 0, measure="T", 
                     showData = TRUE, index = l,
                     showPosteriorSupport = TRUE)
    }
}
dev.off()
```

We'll combine plots and save as PDFs too for easy inserting into documents.
```{r}
# change filename for different non.focalvariable types
filename =  paste("Figures/PresAbs_predictionsALL_byTaxon_type1.pdf") #change if needed
pdf(file = filename, height = 3, width = 15)

for(l in 1:length(sp)){
  par(mfrow = c(1,5))
  for(k in 1:length(covariates)){
    m = models[[1]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate,
                                 non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
    plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                 showData = TRUE, index = l,
                 showPosteriorSupport = FALSE)
}

}

filename =  paste("Figures/LogAbun_predictionsALL_byTaxon_type1.pdf") #change if needed
pdf(file = filename, height = 3, width = 15)

for(l in 1:length(sp)){
  par(mfrow = c(1,5))
  for(k in 1:length(covariates)){
    m = models[[2]]
    covariate = covariates[[k]]
    Gradient = constructGradient(m,focalVariable = covariate,
                                 non.focalVariables = 1)
    predY = predict(m, Gradient=Gradient, expected = TRUE)
    plotGradient(m, Gradient, pred=predY, yshow = 0, measure="Y", 
                 showData = TRUE, index = l,
                 showPosteriorSupport = FALSE)
}

  }
dev.off()
```



#### **Make spatial predictions**

First load the spatial data as a raster stack and check to make sure it looks right.
```{r, echo = T}
library(raster)
#path <- file.path(data.directory, paste0("env-rasters2/backscatter.tif"))
# backscatter
r1 <- raster(file.path(data.directory, paste0("env-rasters2/backscatter.tif")))

# curvature
r2 <- raster(file.path(data.directory, paste0("env-rasters2/curv_07.tif")))

# depth
r3 <- raster(file.path(data.directory, paste0("env-rasters2/depth_01.tif")))

# nTows
r4 <- raster(file.path(data.directory, paste0("env-rasters2/nTows_2020.tif")))

# tsTow
r5 <- raster(file.path(data.directory, paste0("env-rasters2/ysTow_2020.tif")))
r5 <- r5*-1

# stack rasters
rs <- stack(r1,r2,r3,r4,r5)
#rs <- stack(r1,r2,r3,r4)

# remove excess rasters
rm(r1,r2,r3,r4,r5)
#rm(r1,r2,r3,r4)

# plot
plot(rs)
```

Then convert rasters to xy data frame.
```{r, echo = T}
# convert to df
rs.xy <- as.data.frame(rs, xy = TRUE) #raster stack as dataframe
rs.xy$ysTow_2020[rs.xy$nTows_2020 == 0] <- -200 #change cells with no tows to ysTow = 50
rs.xy <- na.omit(rs.xy) # remove NAs
#rs.xy$tsClosure <- 19
XData.grid <- rs.xy[,-c(1,2)] #remove XY columns
summary(XData.grid)

# check for NAs
anyNA(XData.grid )

# save XY to file
xy <- as.matrix(cbind(rs.xy$x, rs.xy$y))
colnames(xy)=c("X","Y")
save(xy, file = "Output/xy.Rdata")
```

**IMPORTANT:** need to make sure any transformations applied in S1 to the covariates is applied here. The names and order of the columns in the 'XData.grid' dataframe should also match those in the Hmsc model object.
```{r, echo = T}
XData.grid$nTows_2020 <- log(XData.grid$nTows_2020+1)
#colnames(XData.grid) <- c('BS','curv','depth','log.nTows','tsClosure')
colnames(XData.grid) <- c('BS','curv','depth','log.nTows','tsTow')
XData.grid <- XData.grid[,c(3,2,1,5,4)]
str(XData.grid)
```

Now we use the prepareGradient function to convert the environmental predictors into a format that can be used by the predict function. We then will predict taxon occurrence (presence/absence) and log abundance conditional on present (abundance COP) to a grid.
```{r, echo = T, warning=FALSE}
# Presence-absence
# prepareGradient
Gradient <- prepareGradient(models[[1]], XDataNew = XData.grid) #this is where you add the interaction grid

# predict taxon occurrence to grid
nParallel = 2 # change nParallel to set number of cores used simultaneously
start <- Sys.time()
# using expected = TRUE to return probabilities, not occurrence
predY <- predict(models[[1]], Gradient=Gradient, expected = TRUE, nParallel= nParallel)
end <- Sys.time()
end - start

# save file
#save(predY, file = "Output/predY_occ_realised.Rdata")

# Log abundance
Gradient <- prepareGradient(models[[2]], XDataNew = XData.grid)
start <- Sys.time()
predY2 <- predict(models[[2]], Gradient=Gradient, expected = TRUE, nParallel = nParallel)  
end <- Sys.time()
end - start
#save(predY2, file = "Output/predY_Abun_realised.Rdata")
```

Using the predicted occurrence grid, we can now map the probability of occurrence and associated uncertainty.

Start by defining the color palettes to be used for the maps.
```{r warning=FALSE}
library(scales)

## hsi colors
### original diverging (temps)
palD <- colour_ramp(c('#009392','#39b185','#9ccb86','#e9e29c','#eeb479','#e88471','#cf597e'))
palD.i <- palD(seq(0,1,length=10))

## abundance colors 
### original geyser diverging
palCG <- colour_ramp(c('#008080','#70a494','#b4c8a8','#f6edbd','#edbb8a','#de8a5a','#ca562c'))
palCG.i <- palCG(seq(0,1,length = 10))
### white,yellow,red continuous
palWYR <- colour_ramp(c('#fdffea', '#F6FF92','#FFC82F','#F16A16','#E7331A'))
palWYR.i <- palWYR(seq(0,1,length=10))

## uncertainty colors
palCB <- colour_ramp(rev(c('#f7feae','#b7e6a5','#7ccba2','#46aea0','#089099','#00718b','#045275')))
palCB.i <- palCB(seq(0,1,length=10))

## hsi colors
### pink-yellow continuous
palCP <- colour_ramp(c('#fef6b5','#ffdd9a','#ffc285','#ffa679','#fa8a76','#f16d7a','#e15383'))
palCP.i <- palCP(seq(0,1,length=10))
### orange-red continuous
palC <- colour_ramp(c('#f6d2a9','#f5b78e','#f19c7c','#ea8171','#dd686c','#ca5268','#b13f64'))
palC.i <- palC(seq(0,1,length = 10))
# blue,yellow,red
palBYR <- colour_ramp(rev(c('#E7331A', '#F16A16', '#FFC82F', '#F6FF92', '#0D98BA', '#0074A5')))
palBYR.i <- palBYR(seq(0,1,length = 10))
# changing the fourth color b/c it's too green
palBYR.i[4]<-'#b7dae2'
show_col(palBYR.i)
```

Load the libraries and define functions needed to generate the distribution maps. 
```{r}
library(ggplot2)
library(colorspace)
library(ggpubr)

# Convert the predY list to an array so we can average
a.PA <- simplify2array(predY) # turn to arrray for averaging
a.A <- simplify2array(predY2)

# Define coefficient of variation (CV) function 
cv <- function(x) abs(sd(x)/mean(x))
```

Plot predicted occurrence and abundance for the year 2020 (19 years post-closure).
```{r, fig.height=7.5,fig.width=5.2, warning=FALSE}
# create contour lines
bathy <- raster(file.path(data.directory, "MORG_sp_ext.tif"))
bathy.xy <- as.data.frame(bathy, xy = TRUE)
bathy.xy <- na.omit(bathy.xy)

# set taxa codes
sp.codes <- read.csv(file = "Data/taxa-codes.csv")

# load taxon occurrence positions
library(sf)
sp.occ <- read_sf("Data/occ.shp")
sp.occ.df <- cbind.data.frame(sp.occ[,c(1:36)],st_coordinates(sp.occ))
sp.occ.df[,c(1:36)][sp.occ.df[,c(1:36)] == 0] <- NA

# load CV results
library(readxl)
auc <- read_excel("Figures/cross_val_Abundance.xlsx",sheet=2)
auc$predAUC <- round(auc$`AUC: Predictive`,digits=2)

r2 <- read_excel("Figures/cross_val_Abundance.xlsx",sheet=3)
r2$predR2 <- round(r2$`R2: Predictive`,digits=2)
```

Make and save prediction plots to PDF file.
```{r, fig.height=7.5,fig.width=5.2, warning=FALSE}
# set filename for pdf to save to
filename =  paste("Figures/Distribution_maps_all.pdf")
pdf(file = filename, height = 7.5, width = 5, compress = FALSE)

for (i in 1:length(sp)){
  MpredY <- apply(a.PA[,i,], 1, mean) # calculate average occurrence probability
  SDpredY <- apply(a.PA[,i,], 1, sd) # calculate SD in occurrence probability (uncertainty)
  
  MpredY2 <- apply(a.A[,i,], 1, mean) # calculate average log abundance
  MpredY2 <- exp(MpredY2) #back transform
  MpredY2 <- MpredY*MpredY2 #hurdle predictions
  
  SDpredY2 <- apply(a.A[,i,], 1, sd) # calculate SD in log abun (uncertainty)
  SDpredY2 <- exp(SDpredY2) #back transform
  SDpredY2 <- SDpredY*SDpredY2 #hurdle SD
  
 
   # convert arrays to dataframes and add the plot data
  Mpred.xyz <- data.frame(x =xy[,1], y =xy[,2], z = MpredY)
  SDpred.xyz <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY)
  Mpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2], z = MpredY2)
  SDpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY2)
  
  # calculate abundance limits
#  binw <- max(Mpred.xyz2$z)/10
 # lims <- c(min(Mpred.xyz2$z), max(Mpred.xyz2$z))
#  brk <- seq(lims[1],lims[2],by=binw)
 # brk.labs <- round(brk, digits =2)
  
  q <- quantile(Mpred.xyz2$z, probs = seq(0.1,0.95, by = 0.05))
#  print(q)
  x95 = as.numeric(q[18])
  brk = seq(0,x95,length.out=11)
  brk = brk[2:10]
  brk.labs <- signif(brk, digits = 2)
  
  # make occurrence df
  sp.occ.i <- sp.occ.df[,c(i,38,39)]
  sp.occ.i <- na.omit(sp.occ.i)
  
  # plot
  a <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
            label=paste0('AUC : ',auc$predAUC[i]), 
            vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz, 
                aes(x=x, y=y, fill=z)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
 #   annotate(geom = 'text', label = paste0(sp.codes$code[i]), x = -Inf, y = Inf, 
  #            hjust = -4.5, vjust = 1.75, size = 5, fill = 'white') +
    scale_fill_stepsn(colors = palBYR.i, breaks = seq(0,1,by=0.1), limits = c(0,1)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'Prob.') +
#    lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  b <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),label=paste0(sp.codes$code[i]), 
             vjust = 1, hjust = 0.25) +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y-200),
             label= paste0('n = ',sum(sp.occ.i[,1])),
             vjust = 1, hjust = 0.25, size = 3) +
    geom_raster(data = SDpred.xyz, 
                aes(x=x, y=y, fill=z)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
    geom_point(data = sp.occ.i, color = 'orange', alpha = 0.5, 
               size = 0.75, fill = 'darkorange',
               aes(x = X, y = Y)) +
    scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD Prob.') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  c <- ggplot() +
     annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
              label=paste0('R2 : ',r2$predR2[i]), 
             vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz2, 
                aes(x=x, y=y, fill=z)) + #change if want to use normal scale
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
  #  scale_fill_stepsn(colors = palBYR.i, limits = c(-0.5,4.5), n.breaks = 10) +
      scale_fill_stepsn(colors = palBYR.i, breaks = brk,
                    limits = c(0,x95), show.limits = FALSE,
                    labels = brk.labs) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'Dens.') +
   # lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.title = element_blank(),
          legend.text = element_text(size = 7.5),
          #legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  d <- ggplot() +
    geom_raster(data = SDpred.xyz2, 
                aes(x=x, y=y, fill=z)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
    scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD Dens.') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  e <- ggarrange(a,b,c,d, labels = c('A','B','C',"D"), 
                 label.x = 0.075, label.y = 0.97)
  print(e)
}

dev.off()
```



Try plotting with gradient color scale.
```{r,fig.height=7.5,fig.width=5.2}
# set filename for pdf to save to
filename =  paste("Figures/Distribution_maps_all_btGr.pdf")
pdf(file = filename, height = 7.5, width = 5, compress = FALSE)

for (i in 1:length(sp)){
  MpredY <- apply(a.PA[,i,], 1, mean) # calculate average occurrence probability
  SDpredY <- apply(a.PA[,i,], 1, sd) # calculate SD in occurrence probability (uncertainty)
  
  MpredY2 <- apply(a.A[,i,], 1, mean) # calculate average log abundance
  MpredY2 <- exp(MpredY2) #back transform
  MpredY2 <- MpredY*MpredY2 #hurdle predictions
 
  SDpredY2 <- apply(a.A[,i,], 1, sd) # calculate SD in log abun (uncertainty)
  SDpredY2 <- SDpredY*SDpredY2 #hurdle SD
  
 
   # convert arrays to dataframes and add the plot data
  Mpred.xyz <- data.frame(x =xy[,1], y =xy[,2], z = MpredY)
  SDpred.xyz <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY)
  Mpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2], z = MpredY2)
  SDpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY2)
  
  # calculate abundance limits
#  binw <- max(Mpred.xyz2$z)/10
 # lims <- c(min(Mpred.xyz2$z), max(Mpred.xyz2$z))
#  brk <- seq(lims[1],lims[2],by=binw)
 # brk.labs <- round(brk, digits =2)
  
  # make occurrence df
  sp.occ.i <- sp.occ.df[,c(i,38,39)]
  sp.occ.i <- na.omit(sp.occ.i) 

  # plot
  a <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
            label=paste0('AUC : ',auc$predAUC[i]), 
            vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz, 
                aes(x=x, y=y, fill=MpredY)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = Band_1)) +
 #   annotate(geom = 'text', label = paste0(sp.codes$code[i]), x = -Inf, y = Inf, 
  #            hjust = -4.5, vjust = 1.75, size = 5, fill = 'white') +
    #scale_fill_stepsn(colors = palD.i, breaks = seq(0,1,by=0.1), limits = c(0,1)) +
    scale_fill_gradient2(low = palD.i[1], mid = palD.i[5], high = palD.i[10],
                         midpoint = max(Mpred.xyz$z)/2,
                         guide = 'colourbar', limits = c(0,1)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'Mean HSI') +
#    lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  b <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),label=paste0(sp.codes$code[i]), 
             vjust = 1, hjust = 0.25) +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y-200),
             label= paste0('n = ',sum(sp.occ.i[,1])),
             vjust = 1, hjust = 0.25, size = 3) +
    geom_raster(data = SDpred.xyz, 
                aes(x=x, y=y, fill=SDpredY)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = Band_1)) +
    geom_point(data = sp.occ.i, color = 'darkred', alpha = 0.5, 
               size = 0.75, fill = 'darkred',
               aes(x = X, y = Y)) +
  #  scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,2)) +
     scale_fill_gradient2(low = palCB.i[1], mid = palCB.i[5], high = palCB.i[10],
                         midpoint = 0.5,
                         guide = 'colourbar', limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD HSI') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  c <- ggplot() +
     annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
              label=paste0('R2 : ',r2$predR2[i]), 
             vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz2, 
                aes(x=x, y=y, fill=MpredY2)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = Band_1)) +
    scale_fill_gradient2(low = palCG.i[1], mid = palCG.i[5], high = palCG.i[10],
                         midpoint = 5, guide = 'colourbar', limits = c(0,10),na.value = palCG.i[10]) +
                     #    oob = scales::squish(x = X, range = c(0,10))) +
  #  scale_fill_stepsn(colors = palCG.i, limits = c(0,5), n.breaks = 10) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = "Dens.") +
   # lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  d <- ggplot() +
    geom_raster(data = SDpred.xyz2, 
                aes(x=x, y=y, fill=SDpredY2)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = Band_1)) +
  #  scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,2)) +
    scale_fill_gradient2(low = palCB.i[1], mid = palCB.i[5], high = palCB.i[10],
                         midpoint = 0.5,
                         guide = 'colourbar', limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD Dens.') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  e <- ggarrange(a,b,c,d, labels = c('A','B','C',"D"), 
                 label.x = 0.075, label.y = 0.97)
  print(e)
 
}

dev.off()
```


Save each plot separately as a PNG for easy embedding in the chapter text.
```{r,warning=FALSE}
# make empty list to store plots
plot_list = list()

# save each plot to list
for (i in 1:length(sp)){
MpredY <- apply(a.PA[,i,], 1, mean) # calculate average occurrence probability
  SDpredY <- apply(a.PA[,i,], 1, sd) # calculate SD in occurrence probability (uncertainty)
  
  MpredY2 <- apply(a.A[,i,], 1, mean) # calculate average log abundance
  MpredY2 <- exp(MpredY2) #back transform
  MpredY2 <- MpredY*MpredY2 #hurdle predictions
  
  SDpredY2 <- apply(a.A[,i,], 1, sd) # calculate SD in log abun (uncertainty)
  SDpredY2 <- exp(SDpredY2) #back transform
  SDpredY2 <- SDpredY*SDpredY2 #hurdle SD
  
 
   # convert arrays to dataframes and add the plot data
  Mpred.xyz <- data.frame(x =xy[,1], y =xy[,2], z = MpredY)
  SDpred.xyz <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY)
  Mpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2], z = MpredY2)
  SDpred.xyz2 <- data.frame(x =xy[,1], y =xy[,2],z = SDpredY2)
  
  # calculate abundance limits
#  binw <- max(Mpred.xyz2$z)/10
 # lims <- c(min(Mpred.xyz2$z), max(Mpred.xyz2$z))
#  brk <- seq(lims[1],lims[2],by=binw)
 # brk.labs <- round(brk, digits =2)
  
  # make occurrence df
  sp.occ.i <- sp.occ.df[,c(i,38,39)]
  sp.occ.i <- na.omit(sp.occ.i)
  
  # plot
  a <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
            label=paste0('AUC : ',auc$predAUC[i]), 
            vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz, 
                aes(x=x, y=y, fill=MpredY)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
 #   annotate(geom = 'text', label = paste0(sp.codes$code[i]), x = -Inf, y = Inf, 
  #            hjust = -4.5, vjust = 1.75, size = 5, fill = 'white') +
    scale_fill_stepsn(colors = palBYR.i, breaks = seq(0,1,by=0.1), limits = c(0,1)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'Prob.') +
#    lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  b <- ggplot() +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),label=paste0(sp.codes$code[i]), 
             vjust = 1, hjust = 0.25) +
    annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y-200),
             label= paste0('n = ',sum(sp.occ.i[,1])),
             vjust = 1, hjust = 0.25, size = 3) +
    geom_raster(data = SDpred.xyz, 
                aes(x=x, y=y, fill=SDpredY)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
    geom_point(data = sp.occ.i, color = 'orange', alpha = 0.5, 
               size = 0.75, fill = 'darkorange',
               aes(x = X, y = Y)) +
    scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD Prob.') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  c <- ggplot() +
     annotate('text',x=max(Mpred.xyz$x),y=max(Mpred.xyz$y),
              label=paste0('R2 : ',r2$predR2[i]), 
             vjust = 1, hjust = 0.5, size = 3) +
    geom_raster(data = Mpred.xyz2, 
                aes(x=x, y=y, fill=log(MpredY2))) + #change if want to use normal scale
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
    scale_fill_stepsn(colors = palBYR.i, limits = c(-0.5,4.5), n.breaks = 10) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'Abun.') +
   # lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.82,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))
  d <- ggplot() +
    geom_raster(data = SDpred.xyz2, 
                aes(x=x, y=y, fill=SDpredY2)) +
    geom_contour(data = bathy.xy, binwidth = 50, colour = 'black', linewidth = 0.25,
                 aes(x = x, y = y, z = MORG_sp_ext)) +
    scale_fill_stepsn(colors = palCB.i, n.breaks = 10, limits = c(0,0.5)) +
    scale_x_continuous(expand = c(0,0), limits = c(6733105,6735505)) +
    scale_y_continuous(expand = c(0,0), limits = c(-3962780,-3959055)) +
    coord_equal() +
    theme_bw() +
    labs(fill = 'SD Log(Ab.)') +
  #  lims(x = c(6733105,6735505), y = c(-3962780,-3959055)) +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(), 
          legend.position = c(0.8,0.22),
          legend.background = element_rect(size = 0.5, colour = 'gray', fill = 'white'),
          legend.text = element_text(size = 7.5),
          legend.title = element_text(size = 9),
          legend.key.width = unit(0.5, 'cm'),
          legend.key.height = unit(0.5, 'cm'),
          panel.border = element_rect(size = 1.5, colour = 'black', fill = NA))

  e <- ggarrange(a,b,c,d, labels = c('A','B','C',"D"), 
                 label.x = 0.075, label.y = 0.97)
  plot_list[[i]] = e
}

# save plots to png (make separate file for each plot)
for (i in 1:length(sp)){
  file_name = paste0('Figures/predicted distribution maps/',sp[i],'_2020-fished.png')
  png(file_name, res = 500, height = 7.5, width = 5, units = 'in')
  print(plot_list[[i]])
  dev.off()
}

rm(a.A, a.PA, plot_list)
```

Map CWM 
```{r}

```
